---
author: Wihan Marais
date: "`r lubridate::today()`"
output: 
  github_document:
    html_preview: FALSE
    toc: TRUE
    toc_depth: 2
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(ragg)
library(lubridate)
library(tidyr)
library(ggplot2)
library(sf)

plot_width <- 6.5
golden_ratio <- (1 + sqrt(5)) / 2

knitr::opts_chunk$set(
  eval = FALSE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  include = TRUE,
  fig.align = "center",
  fig.width = plot_width,
  fig.height = plot_width / golden_ratio,
  out.width = "50%",
  dev = "ragg_png",
  dpi = 180,
  fig.retina = 2,
  cache = FALSE,
  cache.path = "cache/",
  comment = "#>",
  cache.path = "man/cache/README-",
  fig.path = "man/figures/README-"
)
```

# `sasf`

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/sasf)](https://CRAN.R-project.org/package=sasf)
<!-- badges: end -->

The goal of `sasf` is to simplify the process of loading and visualising spatial data for South Africa in `R`.
Shapefiles encompass various administrative levels, such as provinces, districts, municipalities, main places, and subplaces, using Census 2011 demarcations.

The main dataset of interest `subplaces` data frame is embedded in the package. 
`_id` columns represent unique numeric identifiers, while `_name` columns provide descriptive names.
`_mdb` columns present string identifiers corresponding to the demarcations of the Municipal Demarcation Board of South Africa for provinces, districts, and municipalities.

`subplaces` is structured hierarchically on the basis of `_id` values, with the exception of districts.
Consider, for example, the subplace "Wemmershoek":

- `subplace_id`: 167007001
- `mainplace_id`: 167007
- `municipality_id`: 167
- `province_id`: 1

## Installation

<!-- This package requires a working installation of [`sf`](https://github.com/r-spatial/sf#installing). -->

You can install the development version of `sasf` from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("WihanZA/sasf")
```

## Basics

```{r basics, eval = TRUE}
# lazy loading of data
lobstr::mem_used()
library(sasf)
lobstr::mem_used()
invisible(subplaces)
lobstr::mem_used()

# geographic hierarchy
subplaces %>%
  # exclude sfc geometry column
  as.data.frame() %>%
  select(ends_with("_id")) %>%
  pivot_longer(everything()) %>%
  group_by(name) %>%
  summarise(n = n_distinct(value)) %>%
  arrange(desc(n))

# filtering
subplaces %>%
  filter(grepl("stellenbosch", municipality_name, ignore.case = TRUE)) %>%
  head()

# plotting
subplaces %>%
  filter(grepl("stellenbosch", municipality_name, ignore.case = TRUE)) %>%
  ggplot() +
  theme_void() +
  geom_sf(
    aes(fill = subplace_id),
    color = "grey50",
    show.legend = FALSE
  )
```


### Aggregating data with `st_union`

```{r aggregating}
# From subplaces to provinces
provinces <- sasf::subplaces %>%
  group_by(province_id) %>%
  summarise(
    geometry = st_union(geometry),
    .groups = "drop"
  )
```

But `st_union` may take a long time to run, so you can use the pre-aggregated `provinces` data frame instead, as in the example below.

```{r aggregatingplot}
ggplot() +
  theme_void() +
  # use subplace fill
  geom_sf(
    data = sasf::subplaces,
    aes(fill = stringr::str_sub(subplace_id, 6, 9)),
    color = NA,
    show.legend = FALSE
  ) +
  # use province borders
  geom_sf(
    data = sasf::provinces,
    color = "white",
    fill = NA,
    linewidth = 1
  ) +
  scale_fill_viridis_d(option = "magma", begin = 0.15)
```


# Acknowledgements

- The definitions and demarcations used in Census 2011 are detailed in the corresponding *[metadata](https://www.statssa.gov.za/census/census_2011/census_products/Census_2011_Metadata.pdf)*, published by Statistics South Africa (2012).

- The wiki by [konektaz](https://github.com/konektaz) offers a useful summary of the hierarchical structure of spatial layers: *[South Africa Census 2011 spatial metadata](https://github.com/konektaz/shape-files/wiki/South-Africa---Census-2011-spatial-metadata)*

- The original shapefiles were sourced from the [OpenUp Data Portal](https://data.openup.org.za/) at this link: *[Census 2011 Boundaries Subplace Layer](https://data.openup.org.za/dataset/census-2011-boundaries-subplace-layer-qapr-gczi/)*