---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
rm(list = ls())
gc()

library(dplyr)
library(ggplot2)
library(sf)
library(ragg)
# library(rsa.shapefiles)
library(sf)
devtools::load_all()

# remove.packages("rsa.shapefiles")
# install.packages("~/Documents/GitHub/Packages/rsa.shapefiles", type = "source")

knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  eval = TRUE,
  cache = TRUE,
  cache.path = "man/cache/README-",
  fig.path = "man/figures/README-",
  out.width = "50%",
  dpi = 300,
  dev = "ragg_png"
)
```

# `rsa.shapefiles`

<!-- badges: start -->
<!-- badges: end -->

This package simplifies the process of loading and visualising spatial data for South Africa in `R`.
Shapefiles encompass various administrative levels, such as provinces, districts, municipalities, main places, and subplaces, using Census 2011 demarcations.

The main dataset of interest `subplaces` data frame is embedded in the package. 
`_id` columns represent unique numeric identifiers, while `_name` columns provide descriptive names.
`_mdb` columns present string identifiers corresponding to the demarcations of the Municipal Demarcation Board of South Africa for provinces, districts, and municipalities.

`subplaces` is structured hierarchically on the basis of `_id` values, with the exception of districts.
Consider, for example, the subplace "Wemmershoek":

- `subplace_id`: 167007001
- `mainplace_id`: 167007
- `municipality_id`: 167
- `province_id`: 1

## Installation

This package requires a working installation of [sf](https://github.com/r-spatial/sf#installing).

You can install the development version from this Github repository via the [`remotes`](https://github.com/r-lib/remotes#readme) package in R.

```r
# install.packages("remotes")
# install.packages("sf")
remotes::install_github("WihanZA/rsa.shapefiles")
library(rsa.shapefiles)
library(sf)
```

## Examples

### Loading and filtering data

```{r filtering}
rsa.shapefiles::subplaces %>%
  filter(grepl("Stellenbosch", municipality_name)) %>%
  distinct(
    mainplace_name,
    subplace_name
  ) %>%
  head()
```

### Plotting data

```{r plotting}
library(ggplot2)

ggplot() +
  theme_void() +
  geom_sf(data = rsa.shapefiles::subplaces, color = "grey50")
```

### Aggregating data with `st_union`

```{r aggregating, eval = F}
# From subplaces to provinces
provinces <- rsa.shapefiles::subplaces %>%
  group_by(province_id) %>%
  summarise(
    geometry = st_union(geometry),
    .groups = "drop"
  )
```

But `st_union` may take a long time to run, so you can use the pre-aggregated `provinces` data frame instead, as in the example below.

```{r aggregatingplot}
ggplot() +
  theme_void() +
  # use subplace fill
  geom_sf(
    data = rsa.shapefiles::subplaces,
    aes(fill = stringr::str_sub(subplace_id, 6, 9)),
    color = NA,
    show.legend = FALSE
  ) +
  # use province borders
  geom_sf(
    data = rsa.shapefiles::provinces,
    color = "white",
    fill = NA,
    linewidth = 1
  ) +
  scale_fill_viridis_d(option = "magma", begin = 0.15)
```


# Acknowledgements

- The definitions and demarcations used in Census 2011 are detailed in the corresponding *[metadata](https://www.statssa.gov.za/census/census_2011/census_products/Census_2011_Metadata.pdf)*, published by Statistics South Africa (2012).

- The wiki by [konektaz](https://github.com/konektaz) offers a useful summary of the hierarchical structure of spatial layers: *[South Africa Census 2011 spatial metadata](https://github.com/konektaz/shape-files/wiki/South-Africa---Census-2011-spatial-metadata)*

- The original shapefiles were sourced from the [OpenUp Data Portal](https://data.openup.org.za/) at this link: *[Census 2011 Boundaries Subplace Layer](https://data.openup.org.za/dataset/census-2011-boundaries-subplace-layer-qapr-gczi/)*
